### 2018-06-26 latex/discussion.tex
revised discussion

%%% TeX-master: "manuscript"

\section{Discussion}

\subsection{Evolutionary and physiological role of NNMT}

The dynamic interactions among metabolism, signal transduction, and gene regulation is still not very well understood. New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology. We here combine phylogenetic analysis and mathematical modelling of a central metabolic pathway supported by experimental analysis to reveal the likely evolutionary development of the NAD biosynthesis and consumption pathway and the physiological role of the methyltransferase NNMT.

We show that the vitamin B3-degrading enzyme NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis. Due to the fact that NamPRT has a very high affinity for Nam whereas NNMT has a rather low affinity for it, NNMT is able to remove excess Nam that would lead to inhibition of NAD-dependent signalling reactions while maintaining NAD concentrations and even increasing NAD consumption fluxes.

The affinities measured for the human enzymes seem to be optimal and further increase in NAD consumption rates might only enabled by compartmentalisation of NAD biosynthesis.

In contrast to our results, it was proposed that NNMT is present in yeast (\textit{Saccharomyces cerevisiae} \cite{Anderson2003}. The enzyme was later further characterised as methyltransferase of the transcription elongation factors TEF1 and TEF2 but not Nam \cite{Wlodarski2011}.

\subsection{General applicability}

It is of course impossible to extend our conclusions to other metabolic processes without analysing them in more detail, but we have noted that the metabolism of other vitamins, such as pyridoxal (vitamin B6) metabolism, also contains vitamin degrading enzymes \todo{source}, that might thus have similar roles. Degrading enzymes do in general not receive the same attention as biosynthetic enzymes, reflected by the observation that NNMT has only recently been analysed in more detail, besides the much earlier recognition of NNMT as potential marker for some types of cancer (e.g. \cite{Okamura1998}) and routine clinical measurements of urine methyl-Nam concentrations in the context of different diseases (e.g. \cite{Pumpo2001,Delaney2005}). Our analysis show, however, that degrading enzymes do not solely modify substrates for better excretion, but can play a vital role both in human physiology as well as in the evolutionary development of biological processes.

\subsection{Limitations of our analysis and outlook}

Among others, we neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine in our analysis, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism. It has been shown earlier that in turn, NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}. One of the challenges in the analysis of this aspect is the availability of in vivo concentration measurements for SAM and the large amount of reactions using it as substrate. The same holds for the analysis of enzyme acetylation versus deacetylation by sirtuins that would enable prediction about enzyme activation or histone state if available.

>>>>

%%% TeX-master: "manuscript"
% !TeX spellcheck=en_GB
\section{Discussion}

The dynamic interactions among metabolism, signal transduction, and gene regulation is still not very well understood. New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology. We here combined phylogenetic analysis and mathematical modelling the NAD biosynthesis and consumption pathway to better understand the evolution of this pathway and  especially the role of the methyltransferase NNMT. Our analysis reveals that that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis. We show that the presence of NNMT enables higher NAD consumption rates in comparison to NamPRT or NADA alone. It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other NAD requiring processes such as cell growth. We furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}. These finding shed new light on the role of NNMT, which has  earlier  been recognized  as potential marker for some types of cancer (e.g. \cite{Okamura1998}), hinting that increased NNMT expression in cancer might serve to remove  NAM produced by increased NAD-dependent signalling. Further analysis is required to better understand which NAD-signalling reactions are affected. But given a better mechanistic understanding and as routine clinical measurements of urine methyl-Nam (MNam) concentrations are available already, MNam could potentially serve as biomarker in the context of different diseases (e.g\cite{Pumpo2001,Delaney2005}).

In our analysis we furthermore noted that NNMT only provides an advantage as long as the NamPRT affinity for Nam is sufficiently high, this implies that treatment with competitive inhibitors that affect the apparent $K_M$ would have stronger effects on cells that express NNMT compared to cells that don´t express NNMT (see Figure 4E-F). This could be important for cancer treatments as the expression of NNMT is relatively low in normal tissues except liver, whereas several types of cancer overexpress NNMT.

We here neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism. It has been shown earlier that NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}. One of the challenges in the analysis of this aspect is the availability of \textit{in vivo} concentration measurements for SAM and the large amount of reactions using it as substrate.  A mathematical model of this pathway has  been published \cite{PMID14637052}  providing the basis for future studies.

Beside gaining new insights into the role of NNMT for NAD-metabolism, our analyses furthermore provided a suggestion why NADA disappears in vertebrates while it co-exists with NamPRT in bacteria. It seems that the combined enzymatic activity of NNMT and NamPRT is optimised to maintain stable NAD-concentration and high NAD-consumption flux, making NADA obsolete.

\section{Conclusions}

We here highlight the co-evolution of several enzymes of the NAD-pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see Figure 8). This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}. We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site specific positive selection event in NNMT (data not shown). To our knowledge this is the first study that comprehensively demonstrates a functional co-evolution of several enzymes of a pathway.
